### 2024-03-06 tex/examples.tex
[ correction ] finish chapter 7

\begin{lemma}[\AgdaFunction{reify[]}]
  We have a function of type
  $\mathrm{Kripke}\,\V\,\C\,\plr{\cdot} \rightarrowtriangle \C$ for any open
  families $\V$ and $\C$.
\end{lemma}
\begin{proof}
  Let $b : \mathrm{Kripke}\,\V\,\C\,\plr{\cdot}\,\Gamma\,A$.
  We give $b$ the identity renaming (corresponding to no extension of the
  context), yielding
  $b\,1 : \plr{\plr{-} \env\V {\cdot} \wand \plr{-} \sdtstile{}\C A}\,\Gamma$.
  By \cref{thm:construct-env}, we have that $\plr{-} \env\V {\cdot}$ is
  equivalent to $I^*$, and by monoidal closure,
  $\plr{I^* \wand \plr{-} \sdtstile{}\C A}$ is equivalent to
  $\plr{-} \sdtstile{}\C A$, so the result follows.
\end{proof}

>>>>

\begin{lemma}[\AgdaFunction{reify[]}]
  We have a function of type
  $\AgdaFunction{Kripke}\,\V\,\C\,\plr{\cdot} \rightarrowtriangle \C$ for any
  open families $\V$ and $\C$.
\end{lemma}
\begin{proof}
  Let $b : \AgdaFunction{Kripke}\,\V\,\C\,\plr{\cdot}\,\Gamma\,A$.
  We give $b$ the identity renaming (corresponding to no extension of the
  context), yielding
  $b\,1 : \plr{\plr{-} \env\V {\cdot} \wand \plr{-} \sdtstile{}\C A}\,\Gamma$.
  By \cref{thm:construct-env}, we have that $\plr{-} \env\V {\cdot}$ is
  equivalent to $I^*$, and by monoidal closure,
  $\plr{I^* \wand \plr{-} \sdtstile{}\C A}$ is equivalent to
  $\plr{-} \sdtstile{}\C A$, so the result follows.
\end{proof}

### 2024-03-06 tex/semantics.tex
[ correction ] finish chapter 7

Having fixed a universe of syntaxes, in which we can build terms, the next thing
to do is to write recursive functions on terms.
With terms being given by an inductive \AgdaSymbol{data} type definition, they
already come with a recursion and an induction principle.
However, these principles do not handle variable-binding, which we have seen
with the fact that we had to write the \AgdaFunction{bind} helper
function for renaming and substitution in \cref{sec:lrsub}.

@@

A basic property of the universe of syntaxes
is that every syntax supports a functorial action on subterms, realised by a
function \AgdaFunction{map-s}.
Its type says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{$\Psi$} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{$\Psi$}.

@@

The definition
\AgdaFunction{Kripke}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaBound{$\C$}%
\AgdaSpace{}\AgdaBound{$\Delta$} is a kind
of function space that describes a \AgdaBound{$\C$}-value parametrised by
\AgdaBound{$\Delta$}-many additional \AgdaBound{$\V$}-values (all correctly
typed and usage-annotated).
It is used to describe how to go under binders in a
Higher-Order Abstract Syntax style: To go under a binder we must
provide semantic interpretations for all the additional variables.

@@

\AgdaFunction{Wrap}
is a device that turns any type family into an equivalent type family
that is definitionally injective in its indices, which helps with
Agda's type inference.
It turns the type family into a parametrised
record with a single field \AgdaField{get} whose type is the type in
the body of the $\lambda$-abstraction.
For understanding the meaning of
\AgdaFunction{Kripke}, \AgdaFunction{Wrap} can be ignored.

If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
\ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.  That is
to say, the Kripke function is expecting a value for each newly bound
variable, at the multiplicity of its annotation, together with the
resources supporting each of those values. We use the ``magic wand''
function space here to enforce the invariant that the freshly bound
variables have usage annotations that are added to the existing
variables, not shared with them. The use of the
\AgdaFunction{$\Box^r$} modality ensures that we can still use it in
the presence of additional variables introduced by weakening.

@@

\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
  If $\V$ is an open family such that there is a function
  $v : {\sqni} \rightarrowtriangle \V$, we get a function of type
  $\mathrm{Kripke}\,\V\,\C \rightarrowtriangle \mathrm{Scope}\,\C$ for any $\C$.
\end{lemma}
\begin{proof}
  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
  That is, $b$ is a Kripke function yielding $\C$-computations
  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
  $\Gamma, \gr0\delta$.
  Then we give the $\wand$-function an argument in context
  $\gr0\gamma, \Delta$, noting that
  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
  as we wanted from the result.
  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
  We produce this via \cref{thm:ren-env-comp} from an environment
  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
  and a renaming which is the complement to that used on $\Box^r$.
\end{proof}

>>>>

Having fixed a universe of syntaxes, in which we can build terms, the next thing
to do is to write recursive functions on terms.
With terms being given by an inductive \AgdaSymbol{data} type definition, they
already come with a recursion and an induction principle.
However, these principles do not handle variable binding, which we have seen
with the fact that we had to write the \AgdaFunction{bind} helper
function for renaming and substitution in \cref{sec:lrsub}.

@@

A basic property of the universe of syntaxes
is that every syntax supports a functorial action on subterms, realised by a
function \AgdaFunction{map-s}.
Its type says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{$\Psi$} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{$\Psi$}.
The things contained in a layer of syntax are inhabitants of extensible open
families --- either subterms with newly bound variables \AgdaBound{$\Theta$} or
semantic interpretations thereof.
Therefore, function \AgdaBound{f} maps between extensible open families
\AgdaBound{X} and \AgdaBound{Y}, which are given the same newly bound variables
and result type.

@@

The definition
\AgdaFunction{Kripke}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaBound{$\C$}%
\AgdaSpace{}\AgdaBound{$\Delta$} is a kind
of function space that describes a \AgdaBound{$\C$}-value parametrised by
\AgdaBound{$\Delta$}-many additional \AgdaBound{$\V$}-values (all correctly
typed and usage-annotated).
It is used to describe how to go under binders in a
parametric higher-order abstract syntax (PHOAS) style~\citep{Chlipala08}:
To go under a binder we must
provide semantic interpretations for all the additional variables.

@@

I introduced \AgdaRecord{Wrap} in \cref{sec:records} as a device that turns a
unary type family into an equivalent type family that is definitionally
injective in its indices, which helps with Agda's type inference.
It turns the original type family into a parametrised
record with a single field \AgdaField{get}, whose type is the type in
the body of the $\lambda$-abstraction, and a constructor
\AgdaInductiveConstructor{mk} mapping from the original family to the wrapped
family.
The green \AgdaFunction{Wrap} is an extension of the blue \AgdaRecord{Wrap} to
type families of arbitrary arity, made possible thanks to the work of
\citet{Allais19}.
For understanding the meaning of
\AgdaFunction{Kripke}, \AgdaFunction{Wrap} can be ignored.

If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
\ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.  That is
to say, the Kripke function is expecting a value for each newly bound
variable, at the multiplicity of its annotation, together with the
resources supporting each of those values. I use the ``magic wand''
function space here to enforce the invariant that the freshly bound
variables have usage annotations that are added to the existing
variables, not shared with them.
This use of the magic wand goes hand-in-hand with the understanding of
environments as separating conjunctions developed in \cref{sec:lrkits}, and both
are crucial refinements of the previous intuitionistic work.
The use of the
\AgdaFunction{$\Box^r$} modality ensures that we can still use it in
the presence of additional variables introduced by weakening.

@@

\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
  If $\V$ is an open family such that there is a function
  $v : {\sqni} \rightarrowtriangle \V$, we get a function of type
  $\AgdaFunction{Kripke}\,\V\,\C \rightarrowtriangle \AgdaFunction{Scope}\,\C$
  for any $\C$.
\end{lemma}
\begin{proof}
  Let $b : \AgdaFunction{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
  That is, $b$ is a Kripke function yielding $\C$-computations
  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
  $\Gamma, \gr0\delta$.
  Then we give the $\wand$-function an argument in context
  $\gr0\gamma, \Delta$, noting that
  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
  as we wanted from the result.
  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
  We produce this via \cref{thm:ren-env-comp} from an environment
  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
  and a renaming which is the complement to that used on $\Box^r$.
\end{proof}
